QuickSched: Task-based parallelism with dependencies and conflicts

### 2014-11-20 paper/paper.tex
yet another iteration, fixed the BH_times figure, formatted the code correctly.

The particle data is sorted hierarchically, following the
octree structure.
Unlike in many codes, where the leaves store an array of
pointers to the underlying particles, the cells, at all
levels, store only a pointer to the first of their own particles,
and the total number of particles.
The current approach, illustrated in \fig{CellParts} is not
only more compact, it also allows a direct and more cache-efficient access
to the list of particles for any inner node of the tree.
The cost of sorting the particles, with a recursive
partitioning similar to QuickSort \cite{ref:Hoare1962},
is in \oh{N\log N}.

@@

The task-based implementation will consist of three
types of tasks:
\begin{itemize}
    \item {\em Self}-interactions in which all particles
        in a single cell interact with all other particles in the
        same cell,
    \item {\em Particle-particle} pair interactions in which
        all particles in a cell interact with all
        particles in another cell,
    \item {\em Particle-cell} pair interactions in which
        all particles in one cell are interacted with the
        center of mass of all other cells in the tree.
\end{itemize}

These tasks can be created recursively over the cell hierarchy
as shown in the function {\tt make\_tasks} in \fig{MakeTasks}.
The function is called on the root cell with the root cell
and {\tt NULL} as its two cell parameters.
The function recurses as follows:
\begin{itemize}
    \item If called with a single (line~6), split (line~7) cell,
        recurse over all the cell's sub-cells (line~9), and all
        pairs of the cell's sub-cells (line~11),
    \item If called with a single unsplit cell (line~13),
        create a self-interaction task as well as a particle-cell
        task on that cell (line~14),
    \item If called with two non-neighbouring cells (line~21),
        do nothing, as these interactions
        will be computed by the particle-cell task,
    \item If called with two neighbouring cells and both cells
        are split (line~33),
        recurse over all pairs of sub-cells spanning
        both cells (line~37), and
    \item If called with two neighbouring cells
        and one of the cells are not split, create
        a particle-particle pair task over both cells.
\end{itemize}
\noindent where every interaction task additionally locks
the cells on which it operates (lines~16, 25, 30, and 42--43).
In order to prevent generating
a large number of very small tasks, the task generation only recurses
if the cells contain more than a minimum number $n_\mathsf{task}$
of threads each (lines~7 and~34).

The particle-particle pair interaction tasks are implemented
by computing the interactions between all particle pairs spanning
both cells in a double for-loop.
The particle-cell interactions for each leaf node are computed by
traversing the tree recursively starting from the root node:
\begin{itemize}
  \item If called with a node that is a hierarchical parent of
    the leaf node, or with a node that is a direct neighbour of
    a hierarchical parent of the leaf node, recurse over the
    node's sub-cells,
  \item Otherwise, compute the interaction between the leaf node's
    center of mass and all the particles in the leaf node.
\end{itemize}

This task decomposition differs from the traditional tree-walk
in the Barnes-Hut algorithm in that the particle-cell interactions
are grouped per leaf, with each leaf doing its own tree walk.
This approach was chosen to maximize the memory locality
of the particle-cell calculation, as the particles in the leaf,
which are traversed for each particle-cell interaction, are
contiguous in memory, and are thus more likely to remain in the
lowest-level cache.

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with uniformly random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks.
Using the above scheme generated 97\,553 tasks, of which
512 self-interaction tasks, ??? particle-particle interaction
task, and ??? particle-cell interaction tasks.
A total of 141\,840 dependencies were generated, along with
104\,392 locks on 37\,449 resources.

@@

\begin{figure}
    \centerline{\epsfig{file=figures/BH_scaling.pdf,width=0.9\textwidth}}
    \caption{Strong scaling and parallel efficiency of the Barnes-Hut tree-code
        computed over 1\,000\,000 particles.
        Solving the N-Body problem takes 3.3\,s, achieving 60\% parallel
        efficiency, over all 64 cores.
        }
    \label{fig:BHResults}
\end{figure}

Unlike the QR decomposition, the results scale well only to
32 cores, achieving 90\% parallel efficiency, and then
level off for increasing numbers of cores.
This, however, is not a problem of the task-based parallel
algorithm, or of QuickSched, but of the memory bandwidth
of the underlying hardware.
\fig{BHTasks} shows the accumulated cost of each task type and of 
QuickSched over the number of cores.
At 64 cores, the scheduler overheads account for only 4.7\% of
the total computational cost, whereas,
as of 30 cores, the cost of both pair types grow by up to
50\%.
This is most probably due to memory bandwidth restrictions, as
the cost of the self interaction tasks, which do twice as much
computation per memory access, only grow by up to 15\%.

\begin{figure}
    \centerline{\epsfig{file=figures/BH_times.pdf,width=0.8\textwidth}}
    \caption{Accumulated cost of each task type and of the overheads
        associated with {\tt qsched\_gettask}.
        As of $\sim 30$ cores, the cost of both pair interaction task
        types grow by up to 50\%.
        The cost of the self interactions, which entail twice as much
        computation per memory access, grow only by at most 15\%.
        The scheduler overheads make up less than 5\% of the total
        time.
        }
    \label{fig:BHTimes}
\end{figure}

@@

The QuickSched library itself is remarkably simple, consisting of
less than 3\,000 lines of code, including comments.
Both examples, which are distributed with QuickSched,
require less than 1\,000 lines of code each.

@@

There are several possible improvements to QuickSched which
have not been addressed in this paper.
The most obvious of which are the following:
\begin{itemize}
    \item {\em Priorities}: The current implementation of
        QuickSched does not take the resource locks into
        account when selecting tasks in the queues,
    \item {\em Work-stealing}: During work-stealing, the
        queues are probed in a random order although
        the total relative cost of the tasks in the queue,
        as well as the length of their critical paths are
        known,
    \item {\em Costs}: The size of the resources used by
        a task are currently not taken into account when
        assigning it to the queues in {\tt qsched\_enqueue},
        or when approximating the cost of a task.
\end{itemize}

@@

% Acknowledgments
\section*{Acknowledgments}
The author would like to thank Sam Townsend for implementing the
kernels in the task-based QR decomposition as part of his MSc Thesis
at Durham University.
The author also thanks Lydia Heck of the Institute for
Computational Cosmology at Durham University for providing access
to, and expertise on, the COSMA cluster used in the performance
evaluation.
This work was supported by a Durham University Seedcorn Grant.

@@

The algorithms, as described in the previous section, have
all been implemented as part of the Open-Source C-language
QuickSched library.\footnote{\url{http://sourceforge.net/projects/quicksched/}}
This section describes the interface functions and how they
are called, and the following section contains
examples of how QuickSched can be used.

@@

The main functions for setting up the scheduler are:
\begin{itemize}
    \item {\tt void qsched\_init( struct qsched *s , int nr\_queues , int flags )} \\
        Initializes a {\tt qsched} object with the given number of queues.
        The {\tt flags} parameter can be set to any bitwise or combination
        of {\tt qsched\_flag\_none},
        {\tt qsched\_flag\_yield}, and {\tt qsched\_flag\_pthread},
        which are described further below.
        This function must be called before any of the other
        functions are used.
        \vspace{1mm}
    \item {\tt void qsched\_free( struct qsched *s )} \\
        Releases all the memory and other resources allocated by the
        given {\tt qsched} object.
        After this function has been called, the {\tt qsched} will
        need to be re-initialized for reuse.
        \vspace{1mm}
    \item {\tt void qsched\_reset( struct qsched *s )} \\
        Clears the tasks and resources in the given {\tt qsched},
        but does not release the allocated memory or change
        the number of queues.
        \vspace{1mm}
    \item {\tt qsched\_task\_t qsched\_addtask( struct qsched *s , int type , unsigned int flags , void *data , int data\_size , int cost )} \\
        Creates a new task within the given {\tt qsched} and returns
        its handle.
        The {\tt type} and {\tt data} field are copied into the {\tt qsched}
        and passed to the execution function when the {\tt qsched} is run.
        The parameter flags is either {\tt task\_flag\_none} or
        {\tt task\_flag\_virtual}.
        Tasks marked as virtual do not have any action associated with them,
        e.g. they are used only to group or otherwise dependencies, and
        are not passed to the execution function in {\tt qsched\_run}.
        \vspace{1mm}
    \item {\tt qsched\_res\_t qsched\_addres( struct qsched *s , int owner , qsched\_res\_t parent )} \\
        Creates a new resource within the given {\tt qsched} and returns
        its handle.
        The owner field is the initial queue ID to which this resource
        should be assigned, or {\tt qsched\_owner\_none}.
        The {\tt parent} field is the handle of the hierarchical parent of
        the new resource or {\tt qsched\_res\_none} if the resource
        has no hierarchical parent.
        \vspace{1mm}
    \item {\tt void qsched\_addlock( struct qsched *s , qsched\_task\_t t , qsched\_res\_t res )} \\
        Append the resource {\tt res} to the task {\tt t}'s list of
        locks.
        The task {\tt t} will then conflict with any other task that
        also locks the resource {\tt res}, its hierarchical parents, or
        any resource hierarchically below it.
        \vspace{1mm}
    \item {\tt void qsched\_adduse( struct qsched *s , qsched\_task\_t t , qsched\_res\_t res )} \\
        Similar to {\tt qsched\_addlock}, yet the resource is only used and
        is not part of a conflict.
        This information is used when assigning tasks to specific queues.
        \vspace{1mm}
    \item {\tt void qsched\_addunlock( struct qsched *s , qsched\_task\_t ta , qsched\_task\_t tb )} \\
        Appends the task {\tt tb} to the list of tasks that the task {\tt ta}
        unlocks.
        The task {\tt tb} then depends on the task {\tt ta}.
        \vspace{1mm}
    \item {\tt void qsched\_run( struct qsched *s , int nr\_threads , qsched\_funtype fun )} \\
        Executes the tasks in the given {\tt qsched} using {\tt nr\_threads}
        threads via the execution function {\tt fun}, as described in the
        previous section.
        Once a {\tt qsched} has been initialized and filled with
        tasks and resources, it can be run more than once.
        \vspace{1mm}
\end{itemize}

@@

OpenMP has the advantage of being available for most compilers
and also potentially providing some extra platform-specific
scheduling features, e.g.~optimal thread location and/or affinity,
and of integrating seamlessly with other parallel parts of the
user application.
The disadvantage of using OpenMP is that it does not provide
any mechanism for yielding a thread if no tasks are available,
i.e. the main loop in {\tt qsched\_gettask}, described in the
previous section, will spin until a task becomes available.
This may be a problem if other parts of the user application
are running concurrently in the background.
Calling {\tt qsched\_init} with the {\tt qsched\_flag\_yield}
forces the use of {\tt pthreads} and uses conditional variables
to wait for a new task to be enqueued if obtaining a task
from any of the queues fails.
This relinquishes the waiting computational thread for other
processes.

@@

The execution function for these tasks simply calls the appropriate
kernels on the matrix tiles given by the task data:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void exec_fun ( int type , void *data ) {
  int *idata = (int *)data;
  int i = idata[0], j = idata[1], k = idata[2];
  switch ( type ) {
    case tDGEQRF:
      DGEQRF( A[i,j] , ... );
      break;
    case tDLARFT:
      DLARFT( A[i,j] , A[i,i] , ... );
      break;
    case tDTSQRF:
      DTSQRF( A[i,j] , A[j,j]... );
      break;
    case tDSSRFT:
      DSSRFT( A[i,j] , A[i,k] , A[k,j] , ... );
      break;
    default:
      error( "Unknown task type." );
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt A} is the matrix over which the QR
decomposition is executed.

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
void comp_self(struct cell *c) {
  if (c->split)
    for (int j = 0; j < 8; j++) {
      comp_self(c->progeny[j]);
      for (int k = j + 1; k < 8; k++)
        comp_pair(c->progeny[j], c->progeny[k]);
    }
  else
    for (int j = 0; j < c->count; j++)
      for (int k = j + 1; k < c->count; k++)
        interact c->parts[j] and c->parts[k].
}

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
enum { tSELF , tPAIR_PP , tPAIR_PC };
void make_tasks(struct qsched *s, struct cell *ci, struct cell *cj) {
  int j, k;
  qsched_task_t tid;
  struct cell *data[2];
  if (cj == NULL) {
    if (ci->split && ci->count > n_task)
      for (j = 0; j < 8; j++) {
        make_tasks(s, ci->progeny[j], NULL);
        for (k = j + 1; k < 8; k++)
          make_tasks(s, ci->progeny[j], ci->progeny[k]);
      }
    else {
      tid = qsched_addtask(s, tSELF, qsched_flags_none, &ci,
                           sizeof(struct cell *),
                           ci->count * ci->count);
      qsched_addlock(s, tid, ci->res);
      tid = qsched_addtask(s, tPAIR_PC, qsched_flags_none, &ci,
                           sizeof(struct cell *), ci->count);
      qsched_addlock(s, tid, ci->res);
    }
  } else if (ci->split && cj->split &&
             ci->count * cj->count > n_task * n_task)
    for (j = 0; j < 8; j++)
      for (k = 0; k < 8; k++)
        make_tasks(s, ci->progeny[j], cj->progeny[k]);
  else {
    data[0] = ci; data[1] = cj;
    tid = qsched_addtask(s, tPAIR_PP, qsched_flags_none, data,
                         sizeof(struct cell *) * 2,
                         ci->count * cj->count);
    qsched_addlock(s, tid, ci->res);
    qsched_addlock(s, tid, cj->res);
    if (ci->split && cj->split) {
      qsched_addunlock(s, ci->com, tid);
      qsched_addunlock(s, cj->com, tid);
    }
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{C-like pseudo-code for recursive task creation
        for the Barnes-Hut tree-code.}
    \label{fig:MakeTasks}
\end{figure}

The functions for the task themselves are relatively
straight-forward and shown in \fig{BHTasks}, and the
execution function can be written as:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void exec_fun ( int type , void *data ) {
  struct cell **cells = (struct cell **)data;
  switch ( type ) {
    case tSELF:
      comp_self( cells[0] );
      break;
    case tPAIR_PP:
      comp_pair( cells[0] , cells[1] );
      break;
    case tPAIR_PC:
      comp_pair_pc( cells[0] , root );
      break;
    case tCOM:
      comp_com( cells[0] );
      break;
    default:
      error( "Unknown task type." );
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}

>>>>

In our implementation, the particle data is sorted hierarchically,
following the octree structure.
Unlike in many codes, where the leaves store an array of
pointers to the underlying particles, which are not necessarily
contiguous in memory, the cells, at all
levels, store only a pointer to the first of their own particles,
and the total number of particles.
The current approach, illustrated in \fig{CellParts} is not
only more compact, it also allows a direct and more cache-efficient access
to the list of particles for any inner node of the tree.
The cost of sorting the particles, with a recursive
partitioning similar to QuickSort \cite{ref:Hoare1962},
is in \oh{N\log N}.

@@

The task-based implementation consists of three
types of tasks:
\begin{itemize}
    \item {\em Self}-interactions, in which all particles
        in a single cell interact with all other particles in the
        same cell,
    \item {\em Particle-particle} pair interactions, in which
        all particles in a cell interact with all
        particles in another cell,
    \item {\em Particle-cell} pair interactions, in which
        all particles in one cell are interacted with the
        center of mass of all other cells in the tree.
\end{itemize}

These tasks can be created recursively over the cell hierarchy
as shown in the function {\tt make\_tasks} in \fig{MakeTasks}.
The function is called on the root cell with the root cell
and {\tt NULL} as its two cell parameters.
The function recurses as follows (line numbers refer to \fig{MakeTasks}:
\begin{itemize}
    \item If called with a single, split cell (lines~6--7),
        recurse over all the cell's sub-cells, and all
        pairs of the cell's sub-cells (lines~8--11),
    \item If called with a single unsplit cell (line~13),
        create a self-interaction task (line~14) as well as a particle-cell
        task on that cell (line~18),
    \item If called with two neighbouring cells and both cells
        are split (line~22),
        recurse over all pairs of sub-cells spanning
        both cells (lines~24--26), and
    \item If called with two neighbouring cells
        and one of the cells are not split, create
        a particle-particle pair task over both cells (line~29),
    \item If called with two non-neighbouring cells,
        do nothing, as these interactions
        will be computed by the particle-cell task.
\end{itemize}
\noindent where every interaction task additionally locks
the cells on which it operates (lines~17, 20, and 32--33).
In order to prevent generating
a large number of very small tasks, the task generation only recurses
if the cells contain more than a minimum number $n_\mathsf{task}$
of threads each (lines~7 and~23).

As shown in \fig{BHTasks}, the particle-self and particle-particle pair
interaction tasks are implemented
by computing the interactions between all particle pairs spanning
both cells in a double for-loop (lines~9--11 and~22-24 therein).
Some extra logic (lines~2--7 and~15--20) is added to deal with
split cells that did not contain enough particles to warrant the
generation of finer tasks.
The particle-cell interactions for each leaf node are computed by
traversing the tree recursively starting from the root node and:
\begin{itemize}
  \item If called with a node that is a hierarchical parent of
    the leaf node, or with a node that is a direct neighbour of
    a hierarchical parent of the leaf node, recurse over the
    node's sub-cells (lines~29--32),
  \item Otherwise, compute the interaction between the leaf node's
    center of mass and all the particles in the leaf node (lines~33--35).
\end{itemize}

This task decomposition differs from the traditional tree-walk
in the Barnes-Hut algorithm in that the particle-cell interactions
are grouped per leaf, with each leaf doing its own tree walk,
as opposed to doing a tree walk for each individual particle.
This approach was chosen to maximize the memory locality
of the particle-cell calculation, as the particles in the leaf,
which are traversed for each particle-cell interaction, are
contiguous in memory, and are thus more likely to remain in the
lowest-level cache during the entire tree-walk.

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with uniformly random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks.
Using the above scheme generated 97\,553 tasks, of which
512 self-interaction tasks, 5\,068 particle-particle interaction
task, and 32\,768 particle-cell interaction tasks.
A total of 43\,416 locks on 37\,449 resources were generated.

@@

\begin{figure}
    \centerline{\epsfig{file=figures/BH_scaling.pdf,width=0.9\textwidth}}
    \caption{Strong scaling and parallel efficiency of the Barnes-Hut tree-code
        computed over 1\,000\,000 particles.
        Solving the N-Body problem takes 323\,ms, achieving 75\% parallel
        efficiency, over all 64 cores.
        }
    \label{fig:BHResults}
\end{figure}

\begin{figure}
    \centerline{\epsfig{file=figures/tasks_bh_dynamic_64.pdf,width=0.9\textwidth}}
    \caption{Task scheduling of the Barnes-Hut tree-code on 64 cores.
      The red tasks correspond to particle self-interactions, the green
      tasks to the particle-particle pair interactions, and the blue
      tasks to the particle-cell interactions.}
    \label{fig:BHTasksPlot}
\end{figure}

Unlike the QR decomposition, the results scale well only to
32 cores, achieving 90\% parallel efficiency, and then
level off for increasing numbers of cores.
This, however, is not a problem of the task-based parallel
algorithm, or of QuickSched, but of the memory bandwidth
of the underlying hardware.
\fig{BHTimes} shows the accumulated cost of each task type and of 
QuickSched over the number of cores.
At 64 cores, the scheduler overheads account for only $\sim 1$\% of
the total computational cost, whereas,
as of 32 cores, the cost of both pair types grow by up to
40\%.
This is most probably due to memory bandwidth restrictions, as
the cost of the particle-cell interaction tasks, which do significantly more
computation per memory access, only grow by up to 10\%.

\begin{figure}
    \centerline{\epsfig{file=figures/BH_times.pdf,width=0.8\textwidth}}
    \caption{Accumulated cost of each task type and of the overheads
        associated with {\tt qsched\_gettask}, summed over all cores.
        As of 32 cores, the cost of both pair interaction task
        types grow by up to 40\%.
        The cost of the particle-cell interactions, which entail significantly more
        computation per memory access, grow only by at most 10\%.
        The scheduler overheads, i.e.~{\tt qsched\_gettask},
        make up less than 1\% of the total time.}
    \label{fig:BHTimes}
\end{figure}

@@

The QuickSched library itself is remarkably simple, consisting of
less than 3\,000 lines of code, including comments.
Both examples, which are distributed with QuickSched,
require less than 1\,000 lines of code each.
For a more complex, large-scale example of a task-based computation
based on the same algorithms, we refer to \cite{ref:Gonnet2014},
for which the scheduler was originally designed, and from
which QuickSched was back-ported.

@@

There are several possible improvements to QuickSched which
have not been addressed in this paper.
The most obvious of which are the following:
\begin{itemize}
    \item {\em Priorities}: The current implementation of
        QuickSched does not take the resource locks into
        account when selecting tasks in the queues, e.g.~it
        may be advantageous, in some settings, to avoid tasks
        which are involved in too many potential conflicts
        and would therefore restrict the maximum degree of
        parallelism when scheduled.
    \item {\em Work-stealing}: During work-stealing, the
        queues are probed in a random order although
        the total relative cost of the tasks in the queue,
        as well as the length of their critical paths are
        known,
    \item {\em Costs}: The size of the resources used by
        a task are currently not taken into account when
        assigning it to the queues in {\tt qsched\_enqueue},
        or when approximating the cost of a task.
\end{itemize}

@@

% Acknowledgments
\section*{Acknowledgments}
The authors would like to thank Lydia Heck of the Institute for
Computational Cosmology at Durham University for providing access
to, and expertise on, the COSMA cluster used in the performance
evaluation.
This work was supported by a Durham University Seedcorn Grant.

@@

This section describes the QuickSched interface functions and how they
are called.

@@

The main functions for setting up the scheduler are:
\begin{itemize}
    \item {\tt void qsched\_init(struct qsched *s, int nr\_queues, int flags)} \\
        Initializes a {\tt qsched} object with the given number of queues.
        The {\tt flags} parameter can be set to any bitwise or combination
        of {\tt qsched\_flag\_none},
        {\tt qsched\_flag\_yield}, and {\tt qsched\_flag\_pthread},
        which are described further below.
        This function must be called before any of the other
        functions are used.
        \vspace{1mm}
    \item {\tt void qsched\_free(struct qsched *s)} \\
        Releases all the memory and other resources allocated by the
        given {\tt qsched} object.
        After this function has been called, the {\tt qsched} will
        need to be re-initialized for reuse.
        \vspace{1mm}
    \item {\tt void qsched\_reset(struct qsched *s)} \\
        Clears the tasks and resources in the given {\tt qsched},
        but does not release the allocated memory or change
        the number of queues.
        \vspace{1mm}
    \item {\tt qsched\_task\_t qsched\_addtask(struct qsched *s, int type, unsigned int flags, void *data, int data\_size, int cost)} \\
        Creates a new task within the given {\tt qsched} and returns
        its handle.
        The {\tt type} and {\tt data} field are copied into the {\tt qsched}
        and passed to the execution function when the {\tt qsched} is run.
        The parameter flags is either {\tt task\_flag\_none} or
        {\tt task\_flag\_virtual}.
        Tasks marked as virtual do not have any action associated with them,
        e.g. they are used only to group or otherwise dependencies, and
        are not passed to the execution function in {\tt qsched\_run}.
        \vspace{1mm}
    \item {\tt qsched\_res\_t qsched\_addres(struct qsched *s, int owner, qsched\_res\_t parent)} \\
        Creates a new resource within the given {\tt qsched} and returns
        its handle.
        The owner field is the initial queue ID to which this resource
        should be assigned, or {\tt qsched\_owner\_none}.
        The {\tt parent} field is the handle of the hierarchical parent of
        the new resource or {\tt qsched\_res\_none} if the resource
        has no hierarchical parent.
        \vspace{1mm}
    \item {\tt void qsched\_addlock(struct qsched *s, qsched\_task\_t t, qsched\_res\_t res)} \\
        Append the resource {\tt res} to the task {\tt t}'s list of
        locks.
        The task {\tt t} will then conflict with any other task that
        also locks the resource {\tt res}, its hierarchical parents, or
        any resource hierarchically below it.
        \vspace{1mm}
    \item {\tt void qsched\_adduse(struct qsched *s, qsched\_task\_t t, qsched\_res\_t res)} \\
        Similar to {\tt qsched\_addlock}, yet the resource is only used and
        is not part of a conflict.
        This information is used when assigning tasks to specific queues.
        \vspace{1mm}
    \item {\tt void qsched\_addunlock(struct qsched *s, qsched\_task\_t ta, qsched\_task\_t tb)} \\
        Appends the task {\tt tb} to the list of tasks that the task {\tt ta}
        unlocks.
        The task {\tt tb} then depends on the task {\tt ta}.
        \vspace{1mm}
    \item {\tt void qsched\_run(struct qsched *s, int nr\_threads, qsched\_funtype fun)} \\
        Executes the tasks in the given {\tt qsched} using {\tt nr\_threads}
        threads via the execution function {\tt fun}, as described in the
        previous section.
        Once a {\tt qsched} has been initialized and filled with
        tasks and resources, it can be run more than once.
        \vspace{1mm}
\end{itemize}

@@

OpenMP has the advantage of being available for most compilers
and also potentially providing some extra platform-specific
scheduling features, e.g.~optimal thread location and/or affinity,
and of integrating seamlessly with other parallel parts of the
user application.
The disadvantage of using OpenMP is that it does not provide
any mechanism for yielding a thread if no tasks are available,
i.e. the main loop in {\tt qsched\_gettask}, described in the
previous section, will spin until a task becomes available.
This may be a problem if other parts of the user application
are running concurrently in the background outside of QuickSched.
Calling {\tt qsched\_init} with the {\tt qsched\_flag\_yield}
forces the use of {\tt pthreads} and uses conditional variables
to wait for a new task to be enqueued if obtaining a task
from any of the queues fails.
This relinquishes the waiting computational thread for other
processes.

@@

The execution function for these tasks simply calls the appropriate
kernels on the matrix tiles given by the task data:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void exec_fun(int type, void *data) {
  int *idata = (int *)data;
  int i = idata[0], j = idata[1], k = idata[2];
  switch (type) {
    case tDGEQRF:
      DGEQRF(A[i, j], ...);
      break;
    case tDLARFT:
      DLARFT(A[i, j], A[i, i], ...);
      break;
    case tDTSQRF:
      DTSQRF(A[i, j], A[j, j], ...);
      break;
    case tDSSRFT:
      DSSRFT(A[i, j], A[i, k], A[k, j], ...);
      break;
    default:
      error("Unknown task type.");
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt A} is the matrix over which the QR
decomposition is executed.

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
void comp_self(struct cell *c) {
  if (c->split) {
    for (int j = 0; j < 8; j++) {
      comp_self(c->progeny[j]);
      for (int k = j + 1; k < 8; k++)
        comp_pair(c->progeny[j], c->progeny[k]);
    }
  } else {
    for (int j = 0; j < c->count; j++)
      for (int k = j + 1; k < c->count; k++)
        interact c->parts[j] and c->parts[k].
  }
}

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
enum { tSELF, tPAIR_PP, tPAIR_PC };
void make_tasks(struct qsched *s, struct cell *ci, struct cell *cj) {
  int j, k;
  qsched_task_t tid;
  struct cell *data[2];
  if (cj == NULL) {
    if (ci->split && ci->count > n_task) {
      for (j = 0; j < 8; j++) {
        make_tasks(s, ci->progeny[j], NULL);
        for (k = j + 1; k < 8; k++)
          make_tasks(s, ci->progeny[j], ci->progeny[k]);
      }
    } else {
      tid = qsched_addtask(s, tSELF, qsched_flags_none, &ci,
                           sizeof(struct cell *),
                           ci->count * ci->count);
      qsched_addlock(s, tid, ci->res);
      tid = qsched_addtask(s, tPAIR_PC, qsched_flags_none, &ci,
                           sizeof(struct cell *), ci->count);
      qsched_addlock(s, tid, ci->res);
    }
  } else if (ci->split && cj->split &&
             ci->count * cj->count > n_task * n_task) {
    for (j = 0; j < 8; j++)
      for (k = 0; k < 8; k++)
        make_tasks(s, ci->progeny[j], cj->progeny[k]);
  } else {
    data[0] = ci; data[1] = cj;
    tid = qsched_addtask(s, tPAIR_PP, qsched_flags_none, data,
                         sizeof(struct cell *) * 2,
                         ci->count * cj->count);
    qsched_addlock(s, tid, ci->res);
    qsched_addlock(s, tid, cj->res);
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{C-like pseudo-code for recursive task creation
        for the Barnes-Hut tree-code.}
    \label{fig:MakeTasks}
\end{figure}

The functions for the task themselves are relatively
straight-forward and shown in \fig{BHTasks}, and the
execution function can be written as:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void exec_fun(int type, void *data) {
  struct cell **cells = (struct cell **)data;
  switch (type) {
    case tSELF:
      comp_self(cells[0]);
      break;
    case tPAIR_PP:
      comp_pair(cells[0], cells[1]);
      break;
    case tPAIR_PC:
      comp_pair_pc(cells[0], root);
      break;
    case tCOM:
      comp_com(cells[0]);
      break;
    default:
      error("Unknown task type.");
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
